Explaining probabilistic prediction on the simplex with Shapley compositions

### 2023-12-13 paper/main.tex
edit paper

Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express is ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$. The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex. The choice of the basis will be discussed latter in Section \ref{sec:explain}.

@@

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region. Being orthogonal to the virginica class composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}
  \centering
  \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplot.png}
    }
  \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplotsum.png}
  }
  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
  \label{fig:3classes}
\end{figure}

@@

In a four classes examples, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit amoung: 0, 1, 2 and 3. Since they are 64 pixels as features which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for clarity. We again use a SVM classifier with an rbf kernel. The same explaintion analysis as before can be applied here but in a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand of this space is divided in four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextex and the carbon to the center of the tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

\subsubsection{More classes}
\label{sec:more}
...

@@

Some may find the fine analysis of the features contributions in cases with more than four classes, like discussed in Section \ref{sec:more}, tricky. Indeed, this require a careful choose of the sequential binary partition (i.e. the basis) and of the subspace to visualize. However, the Shapley explaination can be summarized by sets of angles, norms and projections.

Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction.

Orthogonality can also be investigated ...
...

@@

>>>>

Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express is ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$. The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex. The choice of the basis will be discussed in Section \ref{sec:moreclasses}.

@@

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region. Being orthogonal to the virginica class composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}
  \centering
  \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplot.png}
    }
  \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
  }
  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
  \label{fig:3classes}
\end{figure}

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity. We again use a SVM classifier with an rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

\subsubsection{More classes}
\label{sec:moreclasses}
When there are more classes, all the dimension of the ILR space cannot be visualized at once but $2$ or $3$-dimensional subspaces can still be visualized. However, in order to choose which ILR components to visualize, one needs to understand what they refer too. The next section provides some intuition about that in the context of the explanation of a classifier prediction with more than 4 classes.

\subsection{Groups of part, balances and sequential binary partition}

....

@@

Some may find the fine analysis of the features contributions in cases with more than four classes tricky. Indeed, this require a careful choose of the sequential binary partition and of the ILR composition to visualize. However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections. Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction. The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory. The projection of a Shapley composition on the set of class composition informs in favor of, or against which, classes a feature is contribution. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation providing the norm of Shapley compositions, angles between them and their projection on class compositions.

@@

\newpage
\section{Summarizing the explanation with norms, angles and projections of Shapley compositions}

TODO...
